Biogeography and Environmental Conditions Shape Phage and Bacteria Interaction Networks Across the Healthy Human Microbiome

### 2017-05-09 doc/manuscript.Rmd
Updated manuscript.

**Word Count**: 237 / 250

@@

Viruses and bacteria are critical components to the human microbiome and play important roles in health and disease. Bacterial communities have been associated with diseases including a wide range of skin conditions [@Hannigan:2013im], acute and chronic wound healing conditions [@Hannigan:2014be; @Loesche:2016ev], and gastrointestinal diseases including inflammatory bowel disease [@He:2016ch; @Norman:2015kb], *Clostridium difficile* infections [@Seekatz:2016fz], and colorectal cancer [@Zackular:2014fba; @Baxter:2014hb]. Altered viromes (virus communities consisting primarily of bacteriophages) have also been associated with various diseases and perturbations, including inflammatory bowel disease [@Norman:2015kb; @Manrique:2016dx], periodontal disease [@Ly:2014ew], spread of antibiotic resistance [@Modi:2013fia], and others [@Monaco:2016ita; @Hannigan:2015fz; @Minot:2011ez; @SantiagoRodriguez:2015gd; @Abeles:2015dy; @Abeles:2014kj]. Viruses interact with bacteria, which together comprise a major portion of total microbial communities [@Haerter:2014ii]. Viruses influence the living microbes of these communities by, for example, lysing their microbial hosts, modulating their functionality through altered host gene expression [@Lindell:2005gz, @Tyler:2013fl, @Hargreaves:2014ja], influencing evolutionary processes through, for example, horizontal gene transfer [@Moon:2015fa, @Modi:2013fi, @Ogg:1981th, @Frost:2005dn] or co-evolution [@Koskella:2014ds], and by altering ecosystem processes and elemental stoichiometry [@Jover:2014gq].

@@

We investigated human bacterial-phage networks by utilizing three published microbiome datasets that contained paired virus and total metagenomic sequence sets, which primarily consisted of bacteria [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei]. Our approach built off of previous large-scale phage-bacteria microbiome network analyses by inferring interactions using metagenomic datasets, instead of using culture-based techniques [@Flores:2013hc]. This metagenomic interaction inference model was powered beyond previous models [@Edwards:2015iz, @Roux:2016cc] by its inclusion of protein interaction data as well as the inclusion of both negative and positive interactions. Through this approach, we were able to provide a basic understanding of the network dynamics associated with healthy human phage and bacterial communities. By building and utilizing a microbiome-virome network, we found that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support communities with distinct communication structures and propensities toward community instability. Through an improved understanding of the healthy state of network structures across the human body, we empower future studies to begin investigating how these community structures are influenced by disease state and their overall impact on human health.

@@

## Cohort Curation and Sample Processing
We studied the differences in virus-bacteria interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets, which explored the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. These datasets were selected because they included virome samples subjected to virus-like particle (VLP) purification. They employed combinations of filtration, chloroform/DNase treatment, and Cesium Chloride treatments to eliminate organismal DNA and thereby allow for direct assessment of both the extracellular and fully-assembled intracellular virome **(Supplemental Figure \ref{SequenceStats} A-B)** [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei]. While the whole metagenomic shotgun sequence samples were not subjected to purification, they primarily consisted of bacteria and had minimal representation from other groups [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei]. Other, minimally represented microbiome members (e.g. fungi, archaea, eukaryotes including *Demodex* mites, etc) were not considered in this study.

Here, the bacterial and viral sequences from these studies were quality filtered and assembled into contigs. We generated clusters of related bacterial and phage contigs based on their k-mer frequencies and co-abundance patterns using the CONCOCT algorithm **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**. Contig lengths within clusters ranged between $10^{3}$ and $10^{5.5}$, confirming the majority of contigs represented genome fragments, rather than complete genomes  **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**. We also confirmed that the majority of bacterial clusters were indeed identified as bacteria (CITE FIGURE HERE), confirming the previous reports [@Minot:2011ez, @Hannigan:2015fz, @Turnbaugh:2009ei]. We interpreted these clusters as operational units of related bacteria and phage genomes, which we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht].

@@

In addition to diet, we also observed an association between obesity and network structure **(Figure \ref{dietnetworks} C-D)**. The obesity-associated network demonstrated a higher degree centrality **(Figure \ref{dietnetworks} C)**, but less closeness centrality compared to the healthy controls **(Figure \ref{dietnetworks} D)**. These results suggest that the obesity network was overall less connected, having microbes further from all other microbes within the community.

@@

Together, our work takes a first step towards defining bacteria-virus interaction profiles as a characteristic of human-associated microbial communities. By focusing on relationships between bacterial and viral communities, we can begin studying them as interacting cohorts they are, instead of isolated entities. This work will also provide a foundation for future studies to begin evaluating how microbiome network dynamics change in disease states, and how the information can be applied therapeutically.

@@

@@

Intuitively, the **degree centrality** of node $V_i$ is defined as the number of edges that are adjacent to that node:

@@

$$ C\left( G \right) =\frac { \sum _{ i=1 }^{ n }{ Cw -c\left( { v }_{ i } \right)  }  }{ { T } }  $$

>>>>

**Word Count**: 242 / 250

@@

Viruses and bacteria are critical components to the human microbiome and play important roles in health and disease. Bacterial communities have been associated with diseases including a wide range of skin conditions [@Hannigan:2013im], acute and chronic wound healing conditions [@Hannigan:2014be; @Loesche:2016ev], and gastrointestinal diseases including inflammatory bowel disease [@He:2016ch; @Norman:2015kb], *Clostridium difficile* infections [@Seekatz:2016fz], and colorectal cancer [@Zackular:2014fba; @Baxter:2014hb]. Altered viromes (virus communities consisting primarily of bacteriophages) have also been associated with various diseases and perturbations including inflammatory bowel disease [@Norman:2015kb; @Manrique:2016dx], periodontal disease [@Ly:2014ew], spread of antibiotic resistance [@Modi:2013fia], and others [@Monaco:2016ita; @Hannigan:2015fz; @Minot:2011ez; @SantiagoRodriguez:2015gd; @Abeles:2015dy; @Abeles:2014kj]. Viruses interact with bacteria, which together comprise a major portion of total microbial communities [@Haerter:2014ii]. Viruses influence the living microbes of these communities by, for example, lysing their microbial hosts, modulating their functionality through altered host gene expression [@Lindell:2005gz, @Tyler:2013fl, @Hargreaves:2014ja], influencing evolutionary processes through, for example, horizontal gene transfer [@Moon:2015fa, @Modi:2013fi, @Ogg:1981th, @Frost:2005dn] or co-evolution [@Koskella:2014ds], and by altering ecosystem processes and elemental stoichiometry [@Jover:2014gq].

@@

We investigated human bacterial-phage networks by utilizing three published microbiome datasets that contained paired virus and total metagenomic sequence sets, which primarily consisted of bacteria [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei]. Our approach built off of previous large-scale phage-bacteria microbiome network analyses by inferring interactions using metagenomic datasets, instead of using culture-based techniques [@Flores:2013hc]. This metagenomic interaction inference model was powered beyond previous models [@Edwards:2015iz, @Roux:2016cc] by its inclusion of protein interaction data as well as the inclusion of both negative and positive interactions. Through this approach, we were able to provide a basic understanding of the network dynamics associated with healthy human phage and bacterial communities. By building and utilizing a microbiome network, we found that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support communities with distinct communication structures and propensities toward community instability. Through an improved understanding of the healthy state of network structures across the human body, we empower future studies to begin investigating how these community structures are influenced by disease state and their overall impact on human health.

@@

## Cohort Curation and Sample Processing
We studied the differences in virus-bacteria interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets, which explored the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. These datasets were selected because they included virome samples subjected to virus-like particle (VLP) purification. To this end, they employed combinations of filtration, chloroform/DNase treatment, and cesium chloride gradients to eliminate organismal DNA and thereby allow for direct assessment of both the extracellular and fully-assembled intracellular virome **(Supplemental Figure \ref{SequenceStats} A-B)** [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei]. While the whole metagenomic shotgun sequence samples were not subjected to purification, they primarily consisted of bacteria [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei].

The bacterial and viral sequences from these studies were quality filtered and assembled into contigs. We generated clusters of related bacterial and phage contigs based on their k-mer frequencies and co-abundance patterns using the CONCOCT algorithm **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**. Contig lengths within clusters ranged between $10^{3}$ and $10^{5.5}$, confirming the majority of contigs represented genome fragments, rather than complete genomes  **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**. We were also able to identify the majority (79%) of bacteria clusters, agreeing with previous reports [@Minot:2011ez, @Hannigan:2015fz, @Turnbaugh:2009ei]. We interpreted these clusters as operational units of related bacteria and phage genomes, which we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht].

@@

In addition to diet, we also observed a potential association between obesity and network structure **(Figure \ref{dietnetworks} C-D)**. The obesity-associated network demonstrated a higher degree centrality **(Figure \ref{dietnetworks} C)**, but less closeness centrality compared to the healthy controls **(Figure \ref{dietnetworks} D)**. These results suggest that the obesity network may have been overall less connected, having microbes further from all other microbes within the community.

@@

Together our work takes an initial step towards defining bacteria-virus interaction profiles as a characteristic of human-associated microbial communities. By focusing on relationships between bacterial and viral communities, we can begin studying them as interacting cohorts they are, instead of isolated entities. This work will also provide a foundation for future studies to begin evaluating how microbiome network dynamics change in disease states, and how the information can be applied therapeutically.

@@

## Bacterial OGU Identification
The longest contig from each OGU was used as its representative sequence for identification. Represetnative sequences were aligned to the European Nucleotide Archive (ENA) bacterial reference database using the blastn algorithm (e-value < $10^{-5}$).

@@

Intuitively, the **degree centrality** of node $V_i$ is defined as the number of edges that are incident to that node:

@@

$$ C\left( G \right) =\frac { \sum _{ i=1 }^{ n }{ Cw -c\left( { V }_{ i } \right)  }  }{ { T } }  $$
